### 2022-11-11 chapter1.tex
Incorporated various clarifications and re-wordings (many suggested by Xiao-Yong)

\section*{Introduction}
Comparative genomics is a powerful tool for yielding insights into evolutionary relationships, molecular function, and the forces that drive gene, genome, and population evolution. These methods often rely on the identification of homologous sequences or homologs, that is sequences with common ancestry, since this ensures that differences between sequences reflect variations in evolution from a common point of divergence. In a sense, each branch that descends from a single ancestral population is a different ``experiment'' by nature, as each is free to explore the evolutionary landscape independent of the others. However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer~\cite{Fitch1970}. The underlying assumption is ``orthologs'' have conserved equivalent functions whereas ``paralogs'', by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor. This technically includes sequences which split by duplication after this point (``in-paralogs''), but excludes sequences which split by duplication before (``out-paralogs'') \cite{Remm2001}. Many current orthology inference pipelines explicitly incorporate steps to detect in-paralogs. However, the resulting orthologs groups can easily be restricted to those without in-paralogs (``single copy orthologs'') for analyses where an assumption of conserved function is necessary.} This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.

@@

Thus, in this work we developed a computational pipeline that can robustly infer and align orthologous groups of proteins even when the genomes are highly redundant (Fig.~\ref{fig:graphical_overview}). We did not seek to reinvent the wheel, so the overall principle of clustering a graph of hits from homology searches is unchanged, though many details were modified to maximize the detection of highly diverged orthologs while also minimizing the impact of incomplete or incorrect annotations. Furthermore, as modern genome annotation pipelines frequently produce gene models and protein sequences in tandem, we implemented an additional clustering step to organize the resulting orthologous groups of proteins into gene-level units. However, most of our efforts were instead focused on the final step of aligning the orthologous sequences. Though genome annotation pipelines are often proficient at identifying the locus of genes, the quality of their gene models can vary considerably and can contain inaccurate exon boundaries or start codons~\cite{Frankish2015, Dunne2018}. As a result, the structure and lengths of the protein sequences that result from these models can be highly heterogeneous, which poses many challenges for their alignment and subsequent analyses.

To develop these methods, we chose a set of 33 assembled and annotated \textit{Drosophila} genomes. All 12 species from the original \textit{Drosophila} 12 Genomes Consortium are included~\cite{D12GC2007}. However, all have been re-sequenced and re-annotated, many with long-read sequencing technology, which has resulted in substantial improvements in the contiguity of their assemblies. Despite this and other genome assembly projects of species in the \textit{Drosophila} genus, there is not yet a database of high-quality alignments of orthologous proteins that reflects these improvements in genome assembly and diversity~\cite{Miller2018, Kim2021}. Given the \textit{Drosophila} genus spans diverse habitats and over 50 million years of evolution but maintains a conserved life cycle and body plan, such a resource would facilitate a new generation of studies that illuminate the forces that drive protein evolution in unprecedented detail~\cite{Wiegmann2011, Obbard2012}.

@@

\section*{Results}
\subsection*{Input genomes}
All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database. The assemblies annotated by the NCBI eukaryotic genome annotation pipeline must past several quality checks and have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}. The \textit{D. miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}. Other annotations were excluded after preliminary clustering showed a deficient in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown). The \textit{D. melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}. The complete list of genomes is given in Table~\ref{stable:genomes}.

Transcripts with alternate UTRs can yield duplicate protein sequences in the annotation. While these alternative UTRs can encode biologically distinct functions, their identical coding sequences make them indistinguishable at the protein level, and the duplicates were therefore removed. Though not strictly necessary, this greatly reduced the computational burden of later steps.

\subsection*{Extraction of best hits from BLAST output}
The protein sequences in each genome annotation were searched against each other in reciprocal pairs using BLAST, yielding a list of high-scoring segment pairs (HSPs) for each query-target pair~\cite{Camacho2009}. HSPs are local alignments, meaning they do not necessarily span the entire lengths of the query and target sequences. Consequently, the search algorithm may return multiple HSPs if statistically significant regions of homology are separated by nonhomologous or poorly conserved regions. Though the most significant HSP is often used to represent all HSPs between a query-target pair, this approach can fail to rank the pairs by their overall significance if their alignments are broken into multiple HSPs. Furthermore, since query-target pairs were later filtered by the amount overlap between their sequences, it can also exclude pairs that pass the overlap threshold even if the most significant HSP alone does not. Thus, HSPs were merged into a single object called a hit. The best hits for each query were then taken as the highest-scoring hits that passed a minimum overlap criterion and were reciprocal between the query and target sequences.

@@

The identification of sets of densely connected nodes in graphs is known as community detection in network analysis. There are many community detection algorithms available, but a popular approach in the context of orthology inference is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster. While this method is robust when clustering smaller or more diverse sets of genomes, the large number of closely related genomes in the input data creates many edges which form triangles but are otherwise poorly supported by the remaining genomes. However, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} = 3 (Fig.~\ref{fig:percolation}). Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships. \textit{k} was set to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).

@@

\subsection*{Grouping orthologous groups by gene}
As genes can have several annotated isoforms, each gene can also be associated with several orthologous groups. However, the orthologous groups are not organized into gene-level units since they were clustered using sequence similarity only. A graph-based approach was therefore used to group orthologous groups with similar sets of parent genes. First, a gene overlap graph was constructed by defining an edge between orthologous groups if the intersection of their associated sets of parent genes is at least 50\% of the smaller of the two. Gene groups were then taken as the connected components of the resulting graph.

\subsection*{Initial alignment and selection of representative sequences}
Since the annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes. As a result, some orthologous genes are annotated with more splice isoforms in some genomes than in others. Under certain conditions, the clustering algorithm can merge smaller or looser clusters of these additional splice isoforms with the main cluster, creating a single orthologous group where some genes have multiple associated sequences. However, these additional sequences can complicate downstream comparative analyses that may not easily generalize to genes with multiple associated sequences. Thus, a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section.

@@

The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}. \textit{a\textsubscript{max}} varies between 0 and 1, but at values above 0.8, the correction severely degrades the alignment of conserved regions. Thus, the sequences were aligned with a conservative value of 0.4. Though some cases of over-alignment remained, most conserved regions were reliably aligned with few to no errors. To correct the remaining errors in alignments of the single copy orthologs, a statistical model of each alignment called a profile HMM was constructed from the well-aligned columns~\cite{Eddy2009}. The original sequences were then aligned to these profiles. By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}). Furthermore, since profile HMMs explicitly model the beginnings and ends of alignments, they can fully exclude these segments from the termini.

@@

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states.” Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. This phylo-HMM was composed of four hidden states, {1A, 1B, 2, 3}, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment (Fig.~\ref{fig:hmm}). Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree. In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments. Fewer than 10\% of alignments were trimmed, but when regions were trimmed, they were largely inferred as state 3 (Fig.~\ref{sfig:hmm_trim}).

@@

For example, despite many additional pre-processing steps and other tweaks introduced by later authors, the basic framework of orthology inference by clustering the hit graph has remained largely unchanged in the past twenty years~\cite{Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}. This longevity is a testament to the robustness of the underlying idea that orthologous proteins should consistently identify each other as the most similar pairs between their genomes. Even as the number of genomes and their taxonomic density has increased dramatically, many orthology inference pipelines continue to use triangle clustering, which was originally applied to sets of far fewer and more distantly related genomes~\cite{Tatusov1997}. This mismatch in scale increases the chance of propagating annotation errors since only a small number of edges are needed to create or merge clusters. Fortunately, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} is a tunable parameter that influences the tightness of a cluster. The optimal value of \textit{k} for a given set of genomes is unclear, but it likely depends on the desired trade-off between sensitivity and specificity. However, percolation theory or simulations may yield insights.

A more subtle difficulty is the annotation of multiple isoforms for a single gene. Though prior pipelines have generally selected the longest isoform as the representative before conducting the orthology searches, we felt this approach would introduce artifacts or other issues during alignment. Instead, as protein sequences are increasingly derived from or linked to genomic sequences, we sought to incorporate the full annotations into the orthology inference pipeline. This, however, created two additional complications. First, a single gene could have several associated orthologous groups if its isoforms belonged to different clusters. Second, a single orthologous group could have several isoforms of a single gene if its isoforms were clustered together. In both cases, the presence of multiple sequences for a single gene creates ambiguities over which is the “primary” isoform. Since the first occurs at the level of orthologous groups, the orthologous groups were first grouped by the similarity of their parent genes using a graph-based strategy. Afterwards, a single representative is easily chosen as the group with the largest number of distinct species, though other criteria are possible. Since the second occurs within an orthologous group, the sequences were first aligned, and a representative for each gene was chosen as the sequence which was most concordant with the initial alignment.

@@

This method corrects most cases of misalignment since poorly supported segments are typically aligned to fewer than four other sequences. However, it does not remove poorly supported segments from columns with more than five non-gap symbols, and their contribution is sometimes sufficient to reproduce the misalignment when the sequences are aligned to the profile HMM. A lightweight convolutional neural network was therefore trained to recognize poorly supported segments to more completely remove them from the profile HMM. The network consisted of a two-dimensional embedding from a 22-dimensional one-hot-encoded vector (20 amino acids, an ambiguous amino acid, and a gap), two convolutional layers, and a one-dimensional output. The first layer was composed of four filters. Two of these filters had a kernel size of 10 and a dilation of 1. The other two filters had a kernel size of 60 with a dilation of 3. The second layer was composed of two filters with a kernel size of 10 and a dilation of 1 (Fig.~\ref{sfig:cnn_architecture}). The network was trained on 88,513 hand-labeled positions across 122 sequences in 46 unique alignments. These examples were largely long segments aligned to few, if any, other sequences (Fig.~\ref{sfig:cnn_training}).

@@

\subsection*{Code and data availability}
The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/orthology_inference2023}. The following Python libraries were used: matplotlib, NumPy, pandas, SciPy, and TensorFlow~\cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020, Abadi2015}. Relevant output files are available in the supporting information. There is no primary data associated with this manuscript. All primary data are available from publicly accessible sources described in their corresponding sections.

>>>>

\section*{Introduction}
Comparative genomics is a powerful tool for yielding insights into evolutionary relationships, molecular function, and the forces that drive gene, genome, and population evolution. These methods often rely on the identification of homologous sequences or homologs, that is sequences with common ancestry, since this ensures that differences between sequences reflect variations in evolution from a common point of divergence. However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer~\cite{Fitch1970}. The underlying assumption is ``orthologs'' have conserved equivalent functions whereas ``paralogs'', by virtue of their redundancy, are more likely to diverge~\cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor. This technically includes sequences which split by duplication after this point (``in-paralogs''), but excludes sequences which split by duplication before (``out-paralogs'') \cite{Remm2001}. Many current orthology inference pipelines explicitly incorporate steps to detect in-paralogs. However, the resulting orthologs groups can easily be restricted to those without in-paralogs (``single copy orthologs'') for analyses where an assumption of conserved function is necessary.} This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.

@@

In this work we therefore developed a computational pipeline that can robustly infer and align orthologous groups of proteins even when the genomes are highly redundant (Fig.~\ref{fig:graphical_overview}). Like many other orthology inference pipelines, our overall approach is based on clustering a graph of hits from homology searches. However, we modified many details to maximize the detection of highly diverged orthologs while also minimizing the impact of incomplete or incorrect annotations. Furthermore, since modern genome annotation pipelines frequently produce gene models and protein sequences in tandem, we implemented an additional clustering step to organize the resulting orthologous groups of proteins into gene-level units. However, most of our efforts were focused on the final step of aligning the orthologous sequences. Though genome annotation pipelines are often proficient at identifying the overall locus of genes, the accurate identification of exon boundaries and start codons when transcript evidence is limited remains an ongoing challenge~\cite{Frankish2015, Dunne2018}. Consequently, protein sequences derived from annotation pipelines can include non-homologous segments of significant length or exclude highly conserved segments. Such heterogeneity in the structure and length of the sequences in an orthologous group poses many challenges for their alignment and subsequent analysis. Thus, we implemented several novel quality control and data cleaning steps to correct mis-alignments and identify likely sequencing, assembly, or annotation errors.

To develop these methods, we chose a set of 33 assembled and annotated \textit{Drosophila} genomes, which includes all 12 species from the original \textit{Drosophila} 12 Genomes Consortium~\cite{D12GC2007}. However, the genomes of these 12 species have been re-sequenced since their first release, which has resulted in substantial improvements in their assemblies and annotations. Despite these developments and other several other recent genome assembly projects of species in the \textit{Drosophila} genus, there is not yet a collection of high-quality alignments of orthologous proteins that reflects these improvements in genome assembly and diversity~\cite{Miller2018, Kim2021}. Given the \textit{Drosophila} genus spans diverse habitats and over 50 million years of evolution but maintains a conserved life cycle and body plan, such a resource would facilitate a new generation of studies that illuminate the forces that drive protein evolution in unprecedented detail~\cite{Wiegmann2011, Obbard2012}.

@@

\section*{Results}
\subsection*{Input genomes and pre-processing}
All annotated genomes in the genus \textit{Drosophila} available in April 2022 were downloaded from NCBI's RefSeq database. The assemblies annotated by the NCBI eukaryotic genome annotation pipeline have passed several quality checks and all have supporting transcript evidence, so the annotations are generally highly complete~\cite{Thibaud-Nissen2013}. The \textit{D. miranda} annotation was excluded due to its unusual karyotype~\cite{Dobzhansky1935}. Other annotations were excluded after preliminary clustering showed a deficiency in the number of orthologous groups containing those genomes, indicating their annotations were less complete (data not shown). The \textit{D. melanogaster} annotation was downloaded from FlyBase~\cite{Gramates2022}. The complete list of genomes is given in Table~\ref{stable:genomes}. Many genes have transcripts that differ only in their UTRs, and as a result there are many duplicate protein sequences in the annotations. Though not strictly necessary, we removed the duplicates in our pipeline, which greatly reduced the computational burden of later steps.

\subsection*{Extraction of best hits from BLAST output}
The protein sequences in each genome annotation were searched against each other in reciprocal pairs using BLAST, yielding a list of high-scoring segment pairs (HSPs) for each query-target pair~\cite{Camacho2009}. HSPs are local alignments, meaning they do not necessarily span the entire lengths of the query and target sequences. Consequently, the search algorithm may return multiple HSPs for each query-target pair if statistically significant regions of homology are separated by nonhomologous or poorly conserved regions. Though the most significant HSP is often used to represent all HSPs between a query-target pair, this approach can fail to rank the pairs by their overall significance if their alignments are broken into multiple HSPs. Furthermore, since query-target pairs were later filtered by the amount overlap between their sequences, it can also exclude pairs that pass the overlap threshold even if the most significant HSP alone does not. Thus, HSPs were merged into a single object called a hit. The best hits for each query were then taken as the highest-scoring hits that passed a minimum overlap criterion and were reciprocal between the query and target sequences.

@@

The identification of sets of densely connected nodes in graphs is known as community detection in network analysis. There are many community detection algorithms available, but a popular approach in the context of orthology inference is building clusters progressively by identifying nodes that form a triangle with at least two other nodes in the cluster. While this method is robust when clustering smaller or more diverse sets of genomes, the large number of closely related genomes in the input data yields many nodes which form triangles but lack dense networks of edges with other nodes. However, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} = 3 (Fig.~\ref{fig:percolation}). Higher values of \textit{k} require greater overlap between a candidate node and those already in the cluster and therefore produce tighter clusters at the cost of excluding more speculative orthology relationships. We set \textit{k} to 4 as compromise between these concerns, yielding 22,813 orthologous groups, a plurality of which contained all 33 species (Fig.~\ref{sfig:clusters}).

@@

\subsection*{Grouping orthologous groups by gene}
As genes can have several annotated isoforms, each gene can be associated with several orthologous groups. However, the orthologous groups are not organized into gene-level units since they were clustered using sequence similarity only. A graph-based approach was therefore used to group orthologous groups with similar sets of parent genes. First, a gene overlap graph was constructed by defining an edge between orthologous groups if the intersection of their associated sets of parent genes is at least 50\% of the smaller of the two. Gene groups were then taken as the connected components of the resulting graph.

\subsection*{Initial alignment and selection of representative sequences}
Since the NCBI annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes. As a result, some genomes are annotated with different or multiple splice isoforms of a given orthologous gene, which can create complex networks in the resulting hit graph. For example, if the genomes are variably annotated with one or both of two distinct isoforms, the resulting graph may contain two clusters connected by a "bridge" formed by the genomes which contain only one of the isoforms. If the nodes bridging the two clusters form cliques with themselves and the clusters, the clique percolation algorithm will merge all the nodes into a single orthologous group where some genes have multiple associated sequences. However, these additional sequences can complicate downstream comparative analyses that may not easily generalize to genes with multiple associated sequences. Thus, in our pipeline a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section.

@@

The aligner MAFFT has a mode for addressing over-alignment with a parameter, \textit{a\textsubscript{max}}, that adjusts the strength of the correction~\cite{Katoh2013, Katoh2016}. \textit{a\textsubscript{max}} varies between 0 and 1, but at values above 0.8, the correction severely degrades the alignment of conserved regions. Thus, in our pipeline the sequences were aligned with a conservative value of 0.4. Though some cases of over-alignment remained, most conserved regions were reliably aligned with few to no errors. To correct the remaining errors in alignments of the single copy orthologs, a statistical model of each alignment called a profile HMM was constructed from the well-aligned columns~\cite{Eddy2009}. The original sequences were then aligned to these profiles. By defining highly conserved “anchor” regions, this approach largely prevents the alignment of chance sequence similarities in long poorly supported segments (Fig.~\ref{fig:realignment}). Furthermore, since profile HMMs explicitly model the beginnings and ends of alignments, they can fully exclude these segments from the termini.

@@

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states” \cite{Felsenstein1996}. Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. This phylo-HMM was composed of four hidden states, {1A, 1B, 2, 3}, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment (Fig.~\ref{fig:hmm}). Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree. In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments. Fewer than 10\% of alignments were trimmed, but when regions were trimmed, they were largely inferred as state 3 (Fig.~\ref{sfig:hmm_trim}).

@@

For example, despite many additional pre-processing steps and other tweaks introduced by later authors, the basic framework of orthology inference by clustering the hit graph has remained largely unchanged in the past twenty years~\cite{Remm2001, Li2003, Jensen2007, Linard2011, Emms2015, Train2017, Cosentino2018}. This longevity is a testament to the robustness of the underlying idea that orthologous proteins should consistently identify each other as the most similar pairs between their genomes. Even as the number of genomes and their taxonomic density has increased dramatically, many orthology inference pipelines continue to use triangle clustering, which was originally applied to sets of far fewer and more distantly related genomes~\cite{Tatusov1997}. This mismatch in scale increases the chance of propagating annotation errors since only a small number of edges are needed to create or merge clusters. Fortunately, triangle clustering is a special case of the more general \textit{k}-clique percolation algorithm where \textit{k} is a tunable parameter that influences the tightness of a cluster. The optimal value of \textit{k} for a given set of genomes is unclear and likely depends on the desired trade-off between sensitivity and specificity. However, percolation theory or simulations may yield insights.

Another challenge is the annotation of multiple isoforms for a single gene. Though prior pipelines have generally selected the longest isoform as the representative before conducting the orthology searches, if the sequences do not share a common intron-exon structure this approach can introduce artifacts or other issues during alignment. Instead, as protein sequences are increasingly derived from or linked to genomic sequences, we sought to incorporate the full annotations into the orthology inference pipeline. This, however, created two additional complications. First, a single gene could have several associated orthologous groups if its isoforms belonged to different clusters. Second, a single orthologous group could have several isoforms of a single gene if its isoforms were clustered together. In both cases, the presence of multiple sequences for a single gene creates ambiguities over which is the “primary” isoform. Since the first occurs at the level of orthologous groups, the orthologous groups were first grouped by the similarity of their parent genes using a graph-based strategy. Afterwards, a single representative is easily chosen as the group with the largest number of distinct species, though other criteria are possible. Since the second occurs within an orthologous group, the sequences were first aligned, and a representative for each gene was chosen as the sequence which was most concordant with this initial alignment.

@@

This method corrects most cases of misalignment since poorly supported segments are typically aligned to fewer than four other sequences. However, it does not remove poorly supported segments from columns with more than five non-gap symbols, and their contribution is sometimes sufficient to reproduce the misalignment when the sequences are aligned to the profile HMM. A lightweight convolutional neural network was therefore trained to recognize poorly supported segments to remove them more completely from the profile HMM. The network consisted of a two-dimensional embedding from a 22-dimensional one-hot-encoded vector (20 amino acids, an ambiguous amino acid, and a gap), two convolutional layers, and a one-dimensional output. The first layer was composed of four filters. Two of these filters had a kernel size of 10 and a dilation of 1. The other two filters had a kernel size of 60 with a dilation of 3. The second layer was composed of two filters with a kernel size of 10 and a dilation of 1 (Fig.~\ref{sfig:cnn_architecture}). The network was trained on 88,513 hand-labeled positions across 122 sequences in 46 unique alignments. These examples were largely long segments aligned to few, if any, other sequences (Fig.~\ref{sfig:cnn_training}).

@@

\subsection*{Code and data availability}
The code used to produce the results and analyses is available at \url{https://github.com/marcsingleton/orthology_inference2023}. HMM algorithms were implemented in the standalone package Homomorph which is available at \url{https://github.com/marcsingleton/homomorph} and on the Python Package Index (PyPI). The following Python libraries were used: matplotlib, NumPy, pandas, SciPy, and TensorFlow~\cite{Hunter2007, Harris2020, McKinney2010, Virtanen2020, Abadi2015}. Relevant output files are available in the supporting information. There is no primary data associated with this manuscript. All primary data are available from publicly accessible sources described in their corresponding sections.
